---
title: "*Crotalus atrox* Sexual Selection Analysis"
output:
  word_document: default
  pdf_document: default
  html_notebook: default
---

#**Overview**
This R Notebook details analyses for a study of sexual selection in the Western diamond-backed rattlesnake (*Crotalus atrox*) sexual selection study (Levine *et al*., in review). Raw data were originally collected, analyzed, and reported by Clark *et al*. 2014.

All data files analyzed and the R Notebook are also stored at https://github.com/brenna-levine/W_Diamondback_Sexual_Selection.

###**1. Data Exploration**
First, I loaded the data file
```{r}
#load csv file
bateman <- read.csv("atrox_bateman_comparison_4feb20.csv")
```
Next, I subset the data to a female set, a male set that included zero mating and reproductive success, and a male set that did not include zero mating and reproductive success. The reason for analyzing non-zero male data is that I (BAL) was worried about the natural dependency of reproductive success on mating success (if mating success = 0, reproductive success has to equal 0). This dependency will be particularly important for calculations of the Bateman gradient that include number of breeding years as a covariate. If number of mates is zero, number of breeding years will also be zero, and therefore we will be basically guaranteed to have a significant interaction between mating success and number of breeding years that will make it impossible to evaluate the effect of number of mates and number of breeding years without considering the interaction between the two. This differs from Levine *et al*. (2015), but given the long-term nature of this data set (with individuals mating over time in multiple years) as opposed to the short-term nature of Levine *et al*. (2015), it's necessary.
```{r}
####subset data
#subset data frame and make new female data frame
female <- subset(bateman, sex=='f')

#subset data frame and make new male data frame
male <- subset(bateman, sex=='m')

#subset new male data frame and make male data frame without zero values
male_nonzero <- subset(male, mates > 0)
```
Next, I plotted the relationships between mating success and reproductive success for the different data sets.


**Female** plot:
```{r}
#create Times New Roman font
windowsFonts(T = "Times New Roman")

#make scatterplot of data
plot(offspring ~ mates, data = female,
	xlab = list("# Mates", cex = 1.1),
	ylab = list("# Offspring", cex = 1.1), 
	main = list("Female Data", cex = 1.1),  xaxt="n",yaxt="n",
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")

#Change x axis to right tic marks, font, and size
axis(1, at=0:6, family = "T", cex.axis=1.2)

#Change y axis to right tic marks, font, and size
axis(2, at=4*0:12, family = "T", cex.axis=1.2)
```
**Male** plot:
```{r}
#make scatterplot of data
plot(offspring ~ mates, data = male,
	xlab = list("# Mates", cex = 1.1),
	ylab = list("# Offspring", cex = 1.1), 
	main = list("Male Data", cex = 1.1),  xaxt="n",yaxt="n",
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")

#Change x axis to right tic marks, font, and size
axis(1, at=0:6, family = "T", cex.axis=1.2)

#Change y axis to right tic marks, font, and size
axis(2, at=4*0:12, family = "T", cex.axis=1.2)
```
**Male plot with data subset to remove zero** value mating success records:
```{r}
#make scatterplot of data
plot(offspring ~ mates, data = male_nonzero,
	xlab = list("# Mates", cex = 1.1),
	ylab = list("# Offspring", cex = 1.1), 
	main = list("Male Data (non-zero)", cex = 1.1),  xaxt="n",yaxt="n",
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")

#Change x axis to right tic marks, font, and size
axis(1, at=0:6, family = "T", cex.axis=1.2)

#Change y axis to right tic marks, font, and size
axis(2, at=4*0:12, family = "T", cex.axis=1.2)
```
I also plotted number of mates vs. number of breeding years for each data set.

**Females**:
```{r}
#make scatterplot of data
plot(mates ~ years_breeding, data = female,
	xlab = list("# Breeding Years", cex = 1.1),
	ylab = list("# Mates", cex = 1.1), 
	main = list("Female Data", cex = 1.1), 
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")
```
**Males**:
```{r}
#make scatterplot of data
plot(mates ~ years_breeding, data = male,
	xlab = list("# Breeding Years", cex = 1.1),
	ylab = list("# Mates", cex = 1.1), 
	main = list("Male Data", cex = 1.1), 
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")
```
**Males without zero values**:
```{r}
#make scatterplot of data
plot(mates ~ years_breeding, data = male_nonzero,
	xlab = list("# Breeding Years", cex = 1.1),
	ylab = list("# Mates", cex = 1.1), 
	main = list("Male Data (non-zero)", cex = 1.1), 
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")
```
I also plotted number of offspring vs number of breeding years for each data set.

**Females**:
```{r}
#make scatterplot of data
plot(offspring ~ years_breeding, data = female,
	xlab = list("# Breeding Years", cex = 1.1),
	ylab = list("# Offspring", cex = 1.1), 
	main = list("Female Data", cex = 1.1), 
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")
```
**Males**:
```{r}
#make scatterplot of data
plot(offspring ~ years_breeding, data = male,
	xlab = list("# Breeding Years", cex = 1.1),
	ylab = list("# Offspring", cex = 1.1), 
	main = list("Male Data", cex = 1.1), 
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")
```
**Males without zero values**:
```{r}
#make scatterplot of data
plot(offspring ~ years_breeding, data = female,
	xlab = list("# Breeding Years", cex = 1.1),
	ylab = list("# Offspring", cex = 1.1), 
	main = list("Male Data (non-zero)", cex = 1.1), 
	family = "T", cex = .8, pch = 1,bty = "l", xaxs = "r")
```


###**2. Bateman Gradients**

####Bateman Gradients--
To test for a significant relationship between number of mates and numbers of offspring (= Bateman gradient, Bss), I ran generalized linear models with a Poisson distribution (because data are count data). This was also done by Apakupakul and Rubenstein (2015).


Starting with the **female** data, I first ran the GLM with an interaction between number of mates and the number of breeding years (a covariate).
```{r}
#glm with interaction between mating success and number of years breeding
glm_female <- glm(offspring ~ mates*years_breeding, data=female, family=poisson)

#view summary of glm
summary(glm_female)
```
There was no evidence of a significant interaction between number of mates and number of breeding years for the female data set, so this interaction was removed and the model re-run.
```{r}
#glm without interaction of fixed effects
#Note: overwriting earlier female glm
glm_female <- glm(offspring ~ mates + years_breeding, data=female, family=poisson)

#view summary of glm
summary(glm_female)
```
Then, I extracted the female Bateman gradient from the model..
```{r}
#print the female Bateman gradient, as extracted from the model coefficients
cat("Female Bss (absolute): ", round(glm_female$coefficients[2], digits = 3), "\n")
```
Then, I tested for a significant effect of number of mates on number of offspring for females by removing mating success from the model with the drop1 function, and comparing the reduced model with a LLR test.
```{r}
#use drop one function with chi-square statistic for LRT
drop1(glm_female, test="Chisq")
```
I found that there was no significant effect of number of mates on reproductive success when adjusting for the number of years that the female bred. This is seen by the LRT which shows that the model is not significantly different if the number of mates is removed as a predictor variable. In other words, **there is no significant Bateman gradient for females when including number of breeding years as a covariate**. This is what we would expect based off of sexual selection theory.

I then ran a similar GLM for **males**, including the zero count data, and with an interaction effect between number of mates and number of breeding years.
```{r}
#glm with interaction between fixed effects
glm_male <- glm(offspring ~ mates*years_breeding, data=male, family=poisson)

#view glm summary
summary(glm_male)
```
As expected given the dependency I describe above (and the large number of zeros in this data set), there was a significant interaction between number of mates and number of breeding years. However, this interaction disappears if I only use the **non-zero male** data for the GLM.
```{r}
#glm of non-zero male data set with interaction between fixed effects
glm_male_nonzero <- glm(offspring ~ mates*years_breeding, data=male_nonzero, family=poisson)

#view summary of glm
summary(glm_male_nonzero)
```
Therefore, I re-ran the male GLM without the interaction term for the non-zero male data (i.e., since there was no significant interaction between number of mates and number of breeding years).
```{r}
#glm without interaction between fixed effects
#Note: overwriting earlier male glm
glm_male_nonzero <- glm(offspring ~ mates + years_breeding, data=male_nonzero, family=poisson)

#view summary of glm
summary(glm_male_nonzero)
```
**Quick sidebar**: I also removed the data files for males with zeros so as to not get confused later.
```{r}
#remove unnecessary data files
rm(male)
rm(glm_male)
```
I extracted the male Bateman gradient from the model coefficients.
```{r}
#cat the male Bateman gradient as extracted from the coefficients of the model
cat("Male Bss (absolute): ", round(glm_male_nonzero$coefficients[2], digits = 3), "\n")
```
I tested the significance of number of mates in the model with drop1, as I did for females.
```{r}
#use drop1 to do LRT with chi-square
drop1(glm_male_nonzero, test="Chisq")
```
Interestingly, I found that there was no significant effect of number of mates on number of offspring for males when adjusting for the number of years that the male bred. In other words, **there is no significant Bateman gradient for males when accounting for this covariate**. This is a really interesting result and supports our results regarding lack of selection on male SVL (see results of SVL analyses later in the notebook). We would expect to see an effect of male SVL on number of offspring produced if males are gaining more priority access to females via male combat (i.e., acquiring more mates). However, we didn't. There is no relationship between male SVL and number of offspring produced (nor number of mates with which offspring were produced). Furthermore, there isn't dramatic sexual size dimorphism in this population, also supporting that male SVL isn't allowing males to gain more offspring via more mating opportunities. Therefore, it makes sense that we wouldn't see a relationship between number of mates (which could have been, but was not, positively related to male SVL) and number of offspring produced.

Then, I evaluated the **residual diagnostics** of the models.

First, for **females**:
```{r}
#plot residual diagnostics for female glm
plot(glm_female)
```
The female residuals look OK considering the sample size is so small (N = 18). There doesn't seem to be a clear pattern on the residuals vs. fitted curve, the qqplot pretty closely follows a straight line, the scale location line is relatively horizontal and the residuals seem randomly spread, and there aren't any points that seem to be influential to the regression results.

Then, I evaluated the residual diagnostics for the **male** GLM.
```{r}
#plot residual diagnostics for male GLM
plot(glm_male_nonzero)
```
Same interpretation of residuals as for females. 

I also **tested for overdispersion** given the Poisson distribution. 

First for **females**:
```{r}
#use package AER to test for overdispersion of female glm
library(AER)
dispersiontest(glm_female)
```
Then for **males**:
```{r}
#use package AER to test for overdispersion of male glm
dispersiontest(glm_male_nonzero) 
```
There is **no evidence of overdispersion for either model**.


####Standardized Bateman Gradients--
All analyses were repeated using **relative mating success and relative reproductive success** so as to quantify standardized Bateman gradients (as in Henshaw *et al*. 2016).

First, I calculated mean mating and reproductive success for each sex. Then, I divided each individual's mating and reproductive success by their respective means to calculate relative mating and reproductive success and added these new values in two new columns to the respective data frame for each sex.
```{r}
###########  females
#calculate mean number of mates for females
mean_mate_f <- mean(female[["mates"]])

#calculate mean number of offspring for females
mean_off_f <- mean(female[["offspring"]])

#cat mean number of mates and offspring for females
cat("Female Mean Mating Success:", round(mean_mate_f, digits = 3), "\n","Female Mean Reproductive Success:", round(mean_off_f, digits = 3), "\n\n")

#add relative mating and reproductive success to female data frame
female$rel_mates <- female$mates/mean_mate_f #rel_mates = relative mating success added to data frame
female$rel_off <- female$offspring/mean_off_f #rel_off = relative reproductive success added to data frame

#view the first few rows of the new data frame to confirm new columns are there
head(female)
cat("\n\n") #cat some new lines

###########  males
#calculate mean number of mates for males
mean_mate_m <- mean(male_nonzero[["mates"]])

#calculate mean number of offspring for males
mean_off_m <- mean(male_nonzero[["offspring"]])

#cat mean number of mates and offspring for males
cat("Male Mean Mating Success:", round(mean_mate_m, digits = 3), "\n", "Male Mean Reproductive Success:", round(mean_off_m, digits = 3), "\n\n")

#add relative mating and reproductive success to male data frame
male_nonzero$rel_mates <- male_nonzero$mates/mean_mate_m #rel_mates = relative mating success added to data frame
male_nonzero$rel_off <- male_nonzero$offspring/mean_off_m #rel_off = relative reproductive success added to data frame

#cat the first few rows of the new data frame to confirm new columns are there
head(male_nonzero)
cat("\n\n") #cat some new lines
```
Starting with the **female** data, I ran the GLM with an interaction between relative number of mates and the number of breeding years. I used a Gaussian probability distribution rather than Poisson, because data are no longer count data and are continuous.
```{r}
#glm with interaction between mating success and number of years breeding
#Note: overwriting the earlier female glm
glm_female <- glm(rel_off ~ rel_mates*years_breeding, data=female, family=gaussian)

#view summary of glm
summary(glm_female)
```
There was no evidence of a significant interaction between number of mates and number of breeding years, so this interaction was removed and the model re-run.
```{r}
#glm without interaction of fixed effects
#Note: overwriting earlier female glm
glm_female <- glm(rel_off ~ rel_mates + years_breeding, data=female, family="gaussian")

#view summary of glm
summary(glm_female)
```
I extracted the standardized female Bss using relative values.
```{r}
#cat the female standardized Bateman gradient as extracted from the model coefficients
cat("Standardized Female Bss (relative): ", round(glm_female$coefficients[2], digits = 3), "\n")
```
Then, I tested for a significant effect of relative number of mates on number of offspring for females by removing relative mating success from the model with the drop1 function, and comparing the reduced model with a LLR test.
```{r}
#use drop one function with chi-square statistic for LRT
drop1(glm_female, test="LRT")
```
**The female standardized Bateman gradient is not significant**.

For **male** data, I ran the GLM with an interaction between relative number of mates and the number of breeding years. I used a Gaussian probability distribution rather than Poisson, because data are no longer count data and are continuous.
```{r}
#glm of non-zero male data set with interaction between fixed effects
#Note: overwriting earlier male glm
glm_male_nonzero <- glm(rel_off ~ rel_mates*years_breeding, data=male_nonzero, family=gaussian)

#view summary of glm
summary(glm_male_nonzero)
```
I re-ran the male GLM without the interaction term for the non-zero male data (since there was no significant interaction between relative number of mates and number of breeding years).
```{r}
#glm without interaction between fixed effects
#Note: overwriting earlier male glm
glm_male_nonzero <- glm(rel_off ~ rel_mates + years_breeding, data=male_nonzero, family=gaussian)

#view summary of glm
summary(glm_male_nonzero)
```
I extracted the standardized Bateman gradient for males from the model coefficients.
```{r}
#cat the male standardized Bateman gradient as extracted from the model coefficients
cat("Standardized Male Bateman Gradient (relative): ", round(glm_male_nonzero$coefficients[2], digits = 3), "\n")
```
I tested the significance of relative number of mates in the model with drop1, as I did for females.
```{r}
#use drop1 to do LRT with chi-square
drop1(glm_male_nonzero, test="Chisq")
```
**The male standardized Bateman gradient is not significant.**

Then, I evaluated the **residual diagnostics** of the models.

First, for **females**:
```{r}
#plot residual diagnostics for female glm
plot(glm_female)
```

Then, for **males**:
```{r}
#plot residual diagnostics for male glm
plot(glm_male_nonzero)
```
The residual diagnostics are similar for both males and females when using the relative, rather than absolute, values in the model with a Gaussian distribution. 

Finally, I made some plots for the publications using ggplot2 and saves them as tiffs.
```{r}
#load graphics packages
library(ggplot2)
library(extrafont)

#save the resulting graph as a tiff
tiff(file="female_Bateman.tiff", width=6, height=4, units="in", res=100)

#plot RRS against StSVL using ggplot for females
ggplot(female, aes(x=mates, y=offspring)) + geom_point(size=2) + theme_classic() + labs(x="Mating Success", y="Reproductive Success") + theme(axis.title.x = element_text(size=14, family = "Arial"), axis.title.y = element_text(size=14, family = "Arial"), axis.text = element_text(size=12, family = "Arial")) 

#over-ride graphics device
dev.off()

#save the resulting graph as a tiff
tiff(file="male_Bateman.tiff", width=6, height=4, units="in", res=100)

#plot RRS against StSVL using ggplot for males
ggplot(male_nonzero, aes(x=mates, y=offspring)) + geom_point(size=2) + theme_classic() + labs(x="Mating Success", y="Reproductive Success") + theme(axis.title.x = element_text(size=14, family = "Arial"), axis.title.y = element_text(size=14, family = "Arial"), axis.text = element_text(size=12, family = "Arial")) 

#over-ride graphics device
dev.off()
```


###**3. Selection on Male SVL**
Next, I tested for selection on male SVL.

First, I loaded the data file and set sire ID as a factor.Then, I calculated relative reproductive success, relative mating success, and mean-standardized SVL from the data.
```{r}
#read raw data
atrox_SVL <- read.csv("atrox_RRS_StSVL.csv")

#make sire ID a factor to later fit as a random effect in the model
atrox_SVL$Sire <- as.factor(atrox_SVL$Sire)

#calculate mean-standardized SVL
mean_svl <- mean(atrox_SVL[["SVL"]]) #calculate mean SVL
stdev_svl <- sd(atrox_SVL[["SVL"]]) #calculate standard deviation SVL
atrox_SVL$StSVL <- (atrox_SVL$SVL - mean_svl)/stdev_svl #calculate mean standardized SVL and add to data frame

#calculate relative reproductive success
mean_RS <- mean(atrox_SVL[["RS"]]) #calculate mean RS
atrox_SVL$RRS <- atrox_SVL$RS/mean_RS #calculate relative RS and add to data frame

#calculate relative mating success
mean_MS <- mean(atrox_SVL[["MS"]]) #calculate mean MS
atrox_SVL$RMS <- atrox_SVL$MS/mean_MS #calculate relative MS and add to data frame

#confirm all calculations added to data frame by viewing the first few rows
head(atrox_SVL) 
```
Then, I ran a linear mixed effects model to test for an effect of mean-standardized annual male SVL on annual reproductive success (= linear selection differential). We have repeated measures in this data set (males siring offspring in multiple years), so I included sire ID as a random effect to account for non-independence among data points.
```{r}
#load package
library(lme4)

#run linear mixed effects model with sire ID set as a random effect
mod_RRS <- lmer(RRS ~ StSVL + (1|Sire), data=atrox_SVL)

#view summary of the model
summary(mod_RRS)
```
Then, I tested the significance of fixed effects in the model (mean-standardized SVL) using the drop1 function and an LRT.
```{r}
#drop1 to test significance of model without StSVL
drop1(mod_RRS, test="Chisq")
```
**There is no significant effect of male mean-standardized SVL on male relative reproductive success**.

I then plotted annual relative reproductive success against mean-standardized SVL.
```{r}
#plot RRS against StSVL using ggplot
ggplot(atrox_SVL, aes(x=StSVL, y=RRS)) + geom_point(size=2) + theme_classic() + labs(x="Standardized SVL", y="Relative Reproductive Success") + theme(axis.title.x = element_text(size=14, family = "Arial"), axis.title.y = element_text(size=14, family = "Arial"), axis.text = element_text(size=12, family = "Arial")) 
```
I also evaluated model residuals.
```{r}
#evalute model residuals
plot(mod_RRS)
```
I also ran a linear mixed effects model to test for an effect of mean-standardized annual male SVL on relative mating success (= linear mating differential). We have repeated measures in this data set (males siring offspring across years), so I included sire ID as a random effect to account for non-independence among data points. *I received a warning* that variance of the sire random effect was very small; this was unsurprising given that almost all male records (except for 3) had the same value of RMS.
```{r}
#use lmer (package lme4) to regress RMS on to StSVL while controlling for repeated measures
mod_RMS <- lmer(RMS ~ StSVL + (1|Sire), data=atrox_SVL)

#view summary of model
summary(mod_RMS)
```
Then, I tested the significance of fixed effects in the model (mean-standardized SVL) using the drop1 function and an LRT.
```{r}
#drop1 to test significance of model without StSVL
drop1(mod_RMS, test="Chisq")
```
**There is not significant effect of male standardized SVL on relative mating success.**

I plotted relative mating success against mean-standardized SVL.
```{r}
#plot RMS against StSVL using ggplot
ggplot(atrox_SVL, aes(x=StSVL, y=RMS)) + geom_point(size=2) + theme_classic() + labs(x="Standardized SVL", y="Relative Mating Success") + theme(axis.title.x = element_text(size=14, family = "Arial"), axis.title.y = element_text(size=14, family = "Arial"), axis.text = element_text(size=12, family = "Arial")) 
```
I also evaluated model residuals. They don't look great, but there's not much to be done about this. This could also be because we have a small sample size and we used a mixed effects model.
```{r}
#plot model residuals
plot(mod_RMS)
```
Finally, I made some plots for publications and saves them as tiffs.
```{r}
#save resulting graph as a tiff
tiff(file="mating_diff.tiff", width=6, height=4, units="in", res=100)

#plot RMS against StSVL using ggplot
ggplot(atrox_SVL, aes(x=StSVL, y=RMS)) + geom_point(size=2) + theme_classic() + labs(x="Mean-Standardized SVL", y="Relative Mating Success") + theme(axis.title.x = element_text(size=14, family = "Arial"), axis.title.y = element_text(size=14, family = "Arial"), axis.text = element_text(size=12, family = "Arial")) 

#over-ride graphics device
dev.off()

#save resulting graph as a tiff
tiff(file="selection_diff.tiff", width=6, height=4, units="in", res=100)

#plot RRS against StSVL using ggplot
ggplot(atrox_SVL, aes(x=StSVL, y=RRS)) + geom_point(size=2) + theme_classic() + labs(x="Mean-Standardized SVL", y="Relative Reproductive Success") + theme(axis.title.x = element_text(size=14, family = "Arial"), axis.title.y = element_text(size=14, family = "Arial"), axis.text = element_text(size=12, family = "Arial")) 

#over-ride graphics device
dev.off()
```


###**4. Opportunities for Sexual Selection and Selection**
Then, I moved on to quantifying opportunities for sexual selection (*I~s~*) and selection (*I*). 

I first quantified *I~s~* and *I* **BY YEAR** for males and females.
To do so, I first read in the data file. Then, I subset the file into a male and a female file.
```{r}
#load csv file
opportunities <- read.csv("opportunities_selection.csv")

####### subset data
#subset data frame and make new female data frame
female_opp <- subset(opportunities, Sex=='f')

#subset data frame and make new male data frame
male_opp <- subset(opportunities, Sex=='m')
```
Starting with the **female** data frame, I automated a subsetting of the female file to create subsets for each year with at least 3 data points. This is because we cannot calculate variance for years in which we have 2 or fewer records.
```{r}
#create a dataframe (c_f) of years for which there are at least 3 female data points
a_f <- rle(sort(female_opp$Year))
b_f <- data.frame(year=a_f$values, n=a_f$lengths)
c_f <- subset(b_f, n>'2') #subset those years that have more than 2 records

#create a subset of the original female data set for each year with greater than 2 records
for(i in c_f$year) { #for each year in the c_f data frame
  
    assign(paste("f_", i, sep = ""), subset(female_opp, Year==i)) #create a subset of the original female data set and name it with an f (for female) and the respective year
  
}

#remove intermediate files
rm(a_f, b_f)
```
For each yearly female data frame, I calculated annual opportunity for sexual selection (*I~s~*).
```{r}
#create list of year data frames (these are the years with >2 records)
df_list <- list(f_2003, f_2004, f_2005, f_2006, f_2007)

#create empty data frame called opp_sex_f in which to store the Is values
opp_sex_f = NULL

#use for loop to calculate variables for female Is for each data frame
for(i in df_list) { #for each data frame in the df_list
  
    variance <- var(i$MS) #calculate the variance in MS
    x <- mean(i$MS) #calculate the mean in MS
    meansq <- x^2 #calculate the squared mean of MS
    rel_var <- variance/meansq #calculate opportunity for sexual selection
    opp_sex_f = rbind(opp_sex_f, data.frame(variance, x, meansq, rel_var)) #add all variables to the new empty data frame
}

#combine the data frame containing years and number of data points for females (c_f) with the data frame containing calculations (opp_sex_f) so that we know what year corresponds wtih each value
opp_sex_f <- cbind(c_f, opp_sex_f)

#### rename columns in opp_sex_f data frame
#load necessary package to do this
library(plyr)

#rename columns of data frame
opp_sex_f <- rename(opp_sex_f, c("year"= "Year", "n"="N (f)", "variance"="Variance MS (f)", "x"="Mean MS (f)", "meansq" = "Squared Mean MS (f)", "rel_var" = "Is (f)"))

#view new data frame
opp_sex_f
```
For each female year data frame, I calculated annual opportunity for selection (*I*).
```{r}
#create empty data frame called opp_sel_f in which to store the I values.
opp_sel_f = NULL

for(i in df_list) { #for each data frame in the previously created df_list
  
    variance <- var(i$RS) #calculate the variance in RS
    x <- mean(i$RS) #calculate the mean in RS
    meansq <- x^2 #calculate the squared mean of RS
    rel_var <- variance/meansq #calculate opportunity for selection
    opp_sel_f = rbind(opp_sel_f, data.frame(variance, x, meansq, rel_var)) #add all variables to the empty data frame

}

#combine the data frame containing years and number of data points for females (c_f) with the data frame containing I calculations (opp_sel_f) so we know which years correspond with which values
opp_sel_f <- cbind(c_f, opp_sel_f)

#rename columns in opp_sel_f
opp_sel_f <- rename(opp_sel_f, c("year"= "Year", "n"="N (f)", "variance"="Variance RS (f)", "x"="Mean RS (f)", "meansq" = "Squared Mean RS (f)", "rel_var" = "I (f)"))

#view new data frame
opp_sel_f
```
I calculated **MEAN** opportunities for sexual selection and selection for **females**.
```{r}
mean_Is_f <- mean(opp_sex_f$`Is (f)`) #mean opportunity for sexual selection
mean_I_f <- mean(opp_sel_f$`I (f)`) #mean opportunity for selection

#cat the means so we know what they are
cat("Female Mean Is: ", round(mean_Is_f, digits = 3), "\n", "Female Mean I: ", round(mean_I_f, digits = 3), "\n\n")
```
Then, for the **male** data frame, I automated a subsetting of the data frame to create subsets for each year with at least 3 data points. This is because we cannot calculate variance for years with fewer than 3 records. This is the same thing I did for females.
```{r}
#create a dataframe (c_m) of years for which there are at least 3 male data points
a_m <- rle(sort(male_opp$Year))
b_m <- data.frame(year=a_m$values, n=a_m$lengths)
c_m <- subset(b_m, n>'2') #subset the male data frame for only years with >2 records

#for each year in the c_m data frame, create a subset of the original male data set
for(i in c_m$year) { #for each year with more than 2 records
  
    assign(paste("m_", i, sep = ""), subset(male_opp, Year==i)) #create a subset for each year for males, and name with an m and the respective year
  
}

#remove intermediate data frames
rm(a_m, b_m)
```
Then, for each annual male data frame, I calculated annual opportunity for sexual selection (*I~s~*).
```{r}
#create list of year data frames (note: overwriting the female df_list)
df_list <- list(m_2003, m_2005, m_2006, m_2007)

#create empty data frame called opp_sex_m in which to store Is values
opp_sex_m = NULL

for(i in df_list) { #for each year data frame in the df_list
  
    variance <- var(i$MS) #calculate the variance in MS
    x <- mean(i$MS) #calculate the mean in MS
    meansq <- x^2 #calculate the squared mean of MS
    rel_var <- variance/meansq #calculate opportunity for sexual selection
    opp_sex_m = rbind(opp_sex_m, data.frame(variance, x, meansq, rel_var)) #add all variables to the empty data frame
}

#combine the data frame containing years and number of data points (c_m) with the data frame containing Is calculations (opp_sex_m) so we know which years correspond with which values
opp_sex_m <- cbind(c_m, opp_sex_m)

#rename columns in opp_sex_m
opp_sex_m <- rename(opp_sex_m, c("year"= "Year", "n"="N (m)", "variance"="Variance MS (m)", "x"="Mean MS (m)", "meansq" = "Squared Mean MS (m)", "rel_var" = "Is (m)"))

#view new data frame
opp_sex_m
```
Then, for each annual male data frame, I calculated annual opportunity for selection (*I*).
```{r}
#create empty data frame in which to store I values
opp_sel_m = NULL

for(i in df_list) { #for each data frame in the df_list
  
    variance <- var(i$RS) #calculate the variance in RS
    x <- mean(i$RS) #calculate the mean in RS
    meansq <- x^2 #calculate the squared mean of RS
    rel_var <- variance/meansq #calculate opportunity for selection
    opp_sel_m = rbind(opp_sel_m, data.frame(variance, x, meansq, rel_var)) #add all variables to the empty data frame
}

#combine the data frame containing years and number of data points (c_m) with the data frame containing I calculations (opp_sel_m) so we know which years correspond with which values
opp_sel_m <- cbind(c_m, opp_sel_m)

#rename columns in opp_sel_m
opp_sel_m <- rename(opp_sel_m, c("year"= "Year", "n"="N (m)", "variance"="Variance RS (m)", "x"="Mean RS (m)", "meansq" = "Squared Mean RS (m)", "rel_var" = "I (m)"))

#view new data frame
opp_sel_m
```
Then, I calculated **MEAN** annual opportunities for sexual selection and selection for **males**.
```{r}
mean_Is_m <- mean(opp_sex_m$`Is (m)`) #mean opportunity for sexual selection
mean_I_m <- mean(opp_sel_m$`I (m)`) #mean opportunity for selection

#cat the means so we know what they are
cat("Male Mean Is: ", round(mean_Is_m, digits = 3), "\n", "Male Mean I: ", round(mean_I_m, digits = 3), "\n\n")
```
I statistically compared **ANNUAL** male and female *I~s~* with an F-ratio test.
```{r}
#create a merged data frame of yearly data that is present for BOTH males and females
opp_sex_merged <- merge(opp_sex_f, opp_sex_m, by.x = "Year") #merge the male and female opp_sex dataframes by year
head(opp_sex_merged) #view the merged data frame to make sure the merge worked correctly
cat("\n\n") #cat some newlines

#write a function to compare male and female Is for each year with F Ratio test
func_sex_comp <- function(x, output) { #function to be applied to each row
  
  #calculate degrees of freedom for males and females
  df_f <- x["N (f)"] - 1 #degrees of freedom for females = #females - 1
  df_m <- x["N (m)"] - 1 #degrees of freedom for males = #males - 1

  #use if/else to determine which values to put in numerator/denominator and order for F Ratio
  if(x["Is (f)"] > x["Is (m)"]){ #if female Is is greater than male Is
  
        crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
        f <- x["Is (f)"]/x["Is (m)"] #use female estimate as numerator in f ratio
        pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
        cat("Note: For", x["Year"],", female Is value is numerator.\n") #cat description
  
  } else { #otherwise
  
        crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
        f <- x["Is (m)"]/x["Is (f)"] #use male estimate as numerator in f ratio
        pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
        cat("Note: For", x["Year"],", male Is value is numerator.\n") #cat description
  
  }
  
  #test whether male and female Is are significantly different by comparing F Ratio with critical value
  if(f > crit){ #if the f ratio is larger than the critical value
  
        cat("For", x["Year"], ", male and female Is are different (F > critical).", "\n", "Female Is: ", round(x["Is (f)"], digits = 3), "\n", "Male Is: ", round(x["Is (m)"], digits = 3), "\n", "F: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Female Degrees of Freedom: ", df_f, "\n", "Male Degrees of Freedom: ", df_m, "\n", "p-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
  } else { #otherwise
  
        cat("For", x["Year"], ", male and female Is are not different (F < critical).", "\n", "Female Is: ", round(x["Is (f)"], digits = 3), "\n", "Male Is: ", round(x["Is (m)"], digits = 3), "\n", "F: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Female Degrees of Freedom: ", df_f, "\n", "Male Degrees of Freedom: ", df_m, "\n", "p-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
  } 
  
}

#run the function on the opp_sex_merged data frame
apply(opp_sex_merged, 1, func_sex_comp)
```
Then, I statistically compared **ANNUAL** male and female *I* with an F-ratio test.
```{r}
#create a merged data frame of yearly data that is present for BOTH males and females
opp_sel_merged <- merge(opp_sel_f, opp_sel_m, by.x = "Year") #merge the male and female opp_sel dataframes by year
head(opp_sel_merged) #view the merged data frame to make sure the merge worked correctly
cat("\n\n") #cat some new lines

#write a function to compare male and female I for each year with an F ratio test
func_sel_comp <- function(x, output) { #function to be applied to each row
  
  #calculate degrees of freedom for males and females
  df_f <- x["N (f)"] - 1 #degrees of freedom for females = #females - 1
  df_m <- x["N (m)"] - 1 #degrees of freedom for males = #males - 1
  
  #use if/else to determine which values to put in numerator/denominator and order for F Ratio
  if(x["I (f)"] > x["I (m)"]){ #if female I is greater than male I
  
        crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
        f <- x["I (f)"]/x["I (m)"] #use female estimate as numerator in f ratio
        pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
        cat("Note: For", x["Year"],", female I value is numerator.\n") #cat description
  
  } else { #otherwise
  
        crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
        f <- x["I (m)"]/x["I (f)"] #use male estimate as numerator in f ratio
        pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
        cat("Note: For", x["Year"],", male I value is numerator.\n") #cat description
  
  }
  
  #test whether male and female I are significantly different by comparing F Ratio with critical value
  if(f > crit){ #if the f ratio is larger than the critical value
  
        cat("For", x["Year"], ", male and female I are different (F > critical).", "\n", "Female I: ", round(x["I (f)"], digits = 3), "\n", "Male I: ", round(x["I (m)"], digits = 3), "\n", "F: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Female Degrees of Freedom: ", df_f, "\n", "Male Degrees of Freedom: ", df_m, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
  } else { #otherwise
  
        cat("For", x["Year"], ", male and female I are not different (F < critical).", "\n", "Female I: ", round(x["I (f)"], digits = 3), "\n", "Male I: ", round(x["I (m)"], digits = 3), "\n", "F: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Female Degrees of Freedom: ", df_f, "\n", "Male Degrees of Freedom: ", df_m, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
  } 
 
}

#run the function on the opp_sel_merged data frame
apply(opp_sel_merged, 1, func_sel_comp)
```
I also statistically compared **MEAN** annual male and female *I~s~* with an F-ratio test.
```{r}
## degrees of freedom for females calculated as 1 - MEAN number of females per year
df_f <- mean(c_f$n)-1 #number of records per year are still stored in the c_f data frame
df_f <- round(df_f, digits = 0) #round to the nearest integer

## degrees of freedom for males calculated as 1 - MEAN number of males per year
df_m <- mean(c_m$n)-1  #number of records per year are still stored in c_m data frame
df_m <- round(df_m, digits = 0) #round to the nearest integer

#calculate F-ratio and critical value for Is, again paying attention to whether males or females have the greater Is value
if(mean_Is_f > mean_Is_m){ #if mean female Is is greater than mean male Is
  
  crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
  f <- mean_Is_f/mean_Is_m #use female estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
  cat("Note: Female Is value is numerator.\n\n") #cat description
  
} else { #otherwise
  
  crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
  f <- mean_Is_m/mean_Is_f #use male estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
  cat("Note: Male Is value is numerator.\n\n") #cat description
  
}

#test whether male and female Is are significantly different by comparing F ratio to critical value
if(f > crit){ #if the f ratio is larger than the critical value
  
  cat("Mean male and female Is are different (F > critical).", "\n", "Male Is: ", round(mean_Is_m, digits = 3), "\n", "Female Is: ", round(mean_Is_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} else { #otherwise
  
  cat("Mean male and female Is are not different (F < critical).", "\n", "Male Is: ", round(mean_Is_m, digits = 3), "\n", "Female Is: ", round(mean_Is_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom:", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n" ) #cat statistical metrics and parameters
  
} 
```
Then, I statistically compared **MEAN** annual male and female *I* with an F-ratio test.
```{r}
#Note: I can use the same degrees of freedom as for Is

#calculate F-ratio and critical value for I
if(mean_I_f > mean_I_m){ #if mean female I is greater than male I
  
  crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
  f <- mean_I_f/mean_I_m #use female estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
  cat("Note: Female I value is numerator.\n\n") #cat description
  
} else { #otherwise
  
  crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
  f <- mean_I_m/mean_I_f #use male estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
  cat("Note: Male I value is numerator.\n\n") #cat description
  
}

#test whether male and female Is are significantly different by comparing F ratio to critical value
if(f > crit){ #if the f ratio is larger than the critical value
  
  cat("Mean male and female I are different (F > critical).", "\n", "Male I: ", round(mean_I_m, digits = 3), "\n", "Female I: ", round(mean_I_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} else { #otherwise
  
  cat("Mean male and female I are not different (F < critical).", "\n", "Male I: ", round(mean_I_m, digits = 3), "\n", "Female I: ", round(mean_I_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} 
```


###**5. Jones Indexes**
Finally, using the means for males and females of annual opportunities for sexual selection and the previously calculated STANDARDIZED Bateman gradients, I calculated the Jones index (*s'~max~* = *B~ss~* x sqrt(*I~s~*)). See Jones (2009). 

I first calculated the Jones index for **females**.
```{r}
#run glm with relative mating and reproductive success and extract coefficients
coef_f <- coef(glm(rel_off ~ rel_mates + years_breeding, data=female, family=gaussian))

#extract Bss from coefficients
Bss_f <- as.numeric(coef_f[2])

#calculate Jones index
Jones_f <- Bss_f*sqrt(mean_Is_f)

#cat the results of the analysis
cat("Female Standardized Bateman Gradient: ", round(Bss_f, digits = 3), "\n","Female Mean Annual Opportunity for Sexual Selection: ", round(mean_Is_f, digits = 3), "\n","Square Root Mean Annual Female Is: ", round(sqrt(mean_Is_f), digits = 3), "\n","Female Jones Index: ", round(Jones_f, digits = 3))
```
Then, I calculated the Jones Index for **males**.
```{r}
#run glm with relative mating and reproductive success and extract coefficients
coef_m <- coef(glm(rel_off ~ rel_mates + years_breeding, data=male_nonzero, family=gaussian))

#extract Bss from coefficients
Bss_m <- as.numeric(coef_m[2])

#calculate Jones index
Jones_m <- Bss_m*sqrt(mean_Is_m)

#cat the results of the analysis
cat("Male Standardized Bateman Gradient: ", round(Bss_m, digits = 3), "\n","Male Mean Annual Opportunity for Sexual Selection: ", round(mean_Is_m, digits = 3), "\n","Square Root Mean Annual Male Is: ", round(sqrt(mean_Is_m), digits = 3), "\n","Male Jones Index: ", round(Jones_m, digits = 3))

```
Finally, I statistically compared the male and female Jones Index (*s'~max~*).
```{r}
# degrees of freedom for females calculated as MEAN number of females per year - 1
df_f <- mean(c_f$n)-1 #number of records per year are still stored in the c_f data frame
df_f <- round(df_f, digits = 0) #round to the nearest integer

## degrees of freedom for males calculated as MEAN number of males per year - 1
df_m <- mean(c_m$n)-1  #number of records per year are still stored in c_m data frame
df_m <- round(df_m, digits = 0) #round to the nearest integer

#calculate F-ratio and critical value for s'max, again paying attention to whether males or females have the greater s'max value
###note: Jones index is in units of SD, so had to convert to variances for F ratio test
if((Jones_f^2) > (Jones_m^2)){ #if squared s'max is greater than squared male s'max
  
  crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
  f <- (Jones_f^2)/(Jones_m^2) #use female estimate (squared) as numerator in f ratio
   pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
  cat("Note: Female s'max (squared) value is numerator.\n\n") #cat description
  
} else { #otherwise
  
  crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
  f <- (Jones_m^2)/(Jones_f^2) #use male estimate (squared) as numerator in f ratio
   pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
  cat("Note: Male s'max (squared) value is numerator.\n\n") #cat description
  
}

#test whether male and female s'max are significantly different by comparing F ratio to critical value
if(f > crit){ #if the f ratio is larger than the critical value
  
  cat("Mean male and female s'max are different (F > critical).", "\n", "Male s'max: ", round(Jones_m, digits = 3), "\n", "Female s'max: ", round(Jones_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} else { #otherwise
  
  cat("Mean male and female s'max are not different (F < critical).", "\n", "Male s'max: ", round(Jones_m, digits = 3), "\n", "Female s'max: ", round(Jones_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom:", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n" ) #cat statistical metrics and parameters
  
} 
```
**There is no significant difference between male and female Jones Indexes.**


###**6. Re-analysis of Copperhead Data**
To have a fair comparison with other pitvipers, I reanalyzed the copperhead (*Agkistrodon contortrix*) data of Levine *et al*. (2015) in the same way. Levine *et al*. (2015) found a significant male Bateman gradient, but included males with zero reproductive success. Here, **I analyzed the data without those zero values** and using the same probability distributions as for the *C. atrox*.

First, I cleared the memory and loaded the copperhead data file.
```{r}
#clear memory
rm(list=ls())

#load csv file
bateman <- read.csv("copperheads.csv")
```
Next, I subset the data frame by males and females, and then I subset the male data frame to retain only those males with non-zero mating and reproductive success. 
```{r}
####subset data
#subset data frame and make new female data frame
female <- subset(bateman, sex=='f')

#subset data frame and make new male data frame
male <- subset(bateman, sex=='m')

#subset new male data frame and make male data frame without zero values
male_nonzero <- subset(male, mates > 0)
```


####Bateman Gradients and Standardized Bateman Gradients for Copperheads--
I ran a glm with a Poisson probability distribution on the male data set without zero values. Note: the copperhead data set did not have males that produced offspring in multiple years, so there is no covariate of number of breeding years in this model.
```{r}
#glm 
glm_male_nonzero <- glm(offspring ~ mates, data=male_nonzero, family=poisson)

#view summary of glm
summary(glm_male_nonzero)
```
Then, I tested for a significant effect of number of mates on number of offspring for males with non-zero mating and reproductive success by removing mating success from the model with the drop1 function, and comparing the reduced model with a LLR test.
```{r}
#use drop one function with chi-square statistic for LRT
drop1(glm_male_nonzero, test="Chisq")
```
**There is a significant effect of mating success on reproductive success for male copperheads** (as was previously found by Levine *et al*. 2015).

Similarly, I also re-ran the female glm for copperheads using the Poisson distribution.
```{r}
#glm
glm_female <- glm(offspring ~ mates, data=female, family=poisson)

#view summary of glm
summary(glm_female)
```
I used the drop1 function to test the significance of mating success on female reproductive success.
```{r}
#use drop one function with chi-square statistic for LRT
drop1(glm_female, test="Chisq")
```
**There is no significant effect of mating success on reproductive success for female copperheads** (as was previously found by Levine *et al*. 2015).

Then, I calculated standardized Bateman gradients for male and female copperheads (using only non-zero data for males) to facilitate calculation of the Jones Index.

First, I had to calculate relative mating and reproductive success for both sexes.
```{r}
###########  females
#calculate mean number of mates for females
mean_mate_f <- mean(female[["mates"]])

#calculate mean number of offspring for females
mean_off_f <- mean(female[["offspring"]])

#add relative mating and reproductive success to female data frame
female$rel_mates <- female$mates/mean_mate_f #rel_mates = relative mating success added to data frame
female$rel_off <- female$offspring/mean_off_f #rel_off = relative reproductive success added to data frame

#view the first few rows of the new data frame to confirm new columns are there
head(female)
cat("\n\n") #cat some new lines

###########  males
#calculate mean number of mates for males
mean_mate_m <- mean(male_nonzero[["mates"]])

#calculate mean number of offspring for males
mean_off_m <- mean(male_nonzero[["offspring"]])

#add relative mating and reproductive success to male data frame
male_nonzero$rel_mates <- male_nonzero$mates/mean_mate_m #rel_mates = relative mating success added to data frame
male_nonzero$rel_off <- male_nonzero$offspring/mean_off_m #rel_off = relative reproductive success added to data frame

#cat the first few rows of the new data frame to confirm new columns are there
head(male_nonzero)
cat("\n\n") #cat some new lines
```
I ren-ran the glm for male copperheads using relative data.
```{r}
#glm
#Note: overwriting the earlier male glm
glm_male_nonzero <- glm(rel_off ~ rel_mates, data=male_nonzero, family=gaussian)

#view summary of glm
summary(glm_male_nonzero)
```
I used the drop1 function to test for a significant effect of relative mating success in the male_nonzero model.
```{r}
#use drop one function with chi-square statistic for LRT
drop1(glm_male_nonzero, test="Chisq")
```
**Male relative mating success is marginally significant in the model (0.05083)**.

I extracted the male standardized Bateman gradient for calculation of the Jones index.
```{r}
#cat the male standardized Bateman gradient as extracted from the model coefficients
cat("Standardized Male Bateman Gradient (relative): ", round(glm_male_nonzero$coefficients[2], digits = 3), "\n")
```
**Male relative mating success has a marginal effect on male reproductive success in male copperheads.**

Next, I ran the glm for females using relative data.
```{r}
#glm
#Note: overwriting the earlier female glm
glm_female <- glm(rel_off ~ rel_mates, data=female, family=gaussian)

#view summary of glm
summary(glm_female)
```
I tested significance of female relative mating success in the model using the drop1 function.
```{r}
#use drop one function with chi-square statistic for LRT
drop1(glm_female, test="Chisq")
```
**Female relative mating success does not have a significant effect on relative reproductive success in copperheads**.

I extracted the standardized Bateman gradient for females.
```{r}
#cat the male standardized Bateman gradient as extracted from the model coefficients
cat("Standardized Female Bateman Gradient (relative): ", round(glm_female$coefficients[2], digits = 3), "\n")
```


####Opportunities for Sexual Selection and Selection in Copperheads--
Then, I compared male and female opportunities for sexual selection and selection, using only males that produced offspring.

First, I calculated male and female *I~s~* and *I* for copperheads.
```{r}
#calculate female opportunity for sexual selection
variance <- var(female$mates) #calculate the variance in MS
x <- mean(female$mates) #calculate the mean in MS
meansq <- x^2 #calculate the squared mean of MS
female_Is <- variance/meansq #calculate opportunity for sexual selection

#calculate female opportunity for selection
variance <- var(female$offspring) #calculate the variance in RS
x <- mean(female$offspring) #calculate the mean in RS
meansq <- x^2 #calculate the squared mean of RS
female_I <- variance/meansq #calculate opportunity for selection

#calculate male opportunity for sexual selection
variance <- var(male_nonzero$mates) #calculate the variance in MS
x <- mean(male_nonzero$mates) #calculate the mean in MS
meansq <- x^2 #calculate the squared mean of MS
male_Is <- variance/meansq #calculate opportunity for sexual selection

#calculate male opportunity for selection
variance <- var(male_nonzero$offspring) #calculate the variance in RS
x <- mean(male_nonzero$offspring) #calculate the mean in RS
meansq <- x^2 #calculate the squared mean of RS
male_I <- variance/meansq #calculate opportunity for selection

```
Then, I calculated degrees of freedom for males and females.
```{r}
#degrees of freedom for males
df_m <- nrow(male_nonzero) - 1

#degrees of freedom for females 
df_f <- nrow(female) - 1

#make degrees of freedom numeric
df_m <- as.numeric(df_m)
df_f <- as.numeric(df_f)
```
I statistically compared male and female *I~s~* for copperheads with an F-ratio test.
```{r}
#calculate F-ratio and critical value for Is, paying attention to whether males or females have the greater Is value
if(female_Is > male_Is){ #if female Is is greater than male Is
  
  crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
  f <- female_Is/male_Is #use female estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
  cat("Note: Female Is value is numerator.\n\n") #cat description
  
} else { #otherwise
  
  crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
  f <- male_Is/female_Is #use male estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
  cat("Note: Male Is value is numerator.\n\n") #cat description
  
}

#test whether male and female Is are significantly different by comparing F ratio to critical value
if(f > crit){ #if the f ratio is larger than the critical value
  
  cat("Mean male and female Is are different (F > critical).", "\n", "Male Is: ", round(male_Is, digits = 3), "\n", "Female Is: ", round(female_Is, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} else { #otherwise
  
  cat("Mean male and female Is are not different (F < critical).", "\n", "Male Is: ", round(male_Is, digits = 3), "\n", "Female Is: ", round(female_Is, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom:", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n" ) #cat statistical metrics and parameters
  
} 
```
Then, I statistically compared male and female *I* for copperheads with an F-ratio test.
```{r}
#calculate F-ratio and critical value for I, paying attention to whether males or females have the greater Is value
if(female_I > male_I){ #if female I is greater than male I
  
  crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
  f <- female_I/male_I #use female estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
  cat("Note: Female I value is numerator.\n\n") #cat description
  
} else { #otherwise
  
  crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
  f <- male_I/female_I #use male estimate as numerator in f ratio
   pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
  cat("Note: Male I value is numerator.\n\n") #cat description
  
}

#test whether male and female I are significantly different by comparing F ratio to critical value
if(f > crit){ #if the f ratio is larger than the critical value
  
  cat("Mean male and female I are different (F > critical).", "\n", "Male I: ", round(male_I, digits = 3), "\n", "Female I: ", round(female_I, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} else { #otherwise
  
  cat("Mean male and female I are not different (F < critical).", "\n", "Male I: ", round(male_I, digits = 3), "\n", "Female I: ", round(female_I, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom:", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n" ) #cat statistical metrics and parameters
  
} 
```


####Jones Indexes for Copperheads--
Finally, I calculated the Jones Index (*s'~max~*) for male and female copperheads, using only males that produced offspring.

First, I calculated the Jones Index for males.
```{r}
#run glm with relative mating and reproductive success and extract coefficients
coef_m <- coef(glm(rel_off ~ rel_mates, data=male_nonzero, family=gaussian))

#extract Bss from coefficients
Bss_m <- as.numeric(coef_m[2])

#calculate Jones index
Jones_m <- Bss_m*sqrt(male_Is)

#cat the results of the analysis
cat("Male Standardized Bateman Gradient: ", round(Bss_m, digits = 3), "\n","Male Opportunity for Sexual Selection: ", round(male_Is, digits = 3), "\n","Square Root Male Is: ", round(sqrt(male_Is), digits = 3), "\n","Male Jones Index: ", round(Jones_m, digits = 3))

```
Then, I calculated the Jones Index for females.
```{r}
#run glm with relative mating and reproductive success and extract coefficients
coef_f <- coef(glm(rel_off ~ rel_mates, data=female, family=gaussian))

#extract Bss from coefficients
Bss_f <- as.numeric(coef_f[2])

#calculate Jones index
Jones_f <- Bss_f*sqrt(female_Is)

#cat the results of the analysis
cat("Female Standardized Bateman Gradient: ", round(Bss_f, digits = 3), "\n","Female Opportunity for Sexual Selection: ", round(female_Is, digits = 3), "\n","Square Root Female Is: ", round(sqrt(female_Is), digits = 3), "\n","Female Jones Index: ", round(Jones_f, digits = 3))

```
Finally, I statistically compared the male and female Jones Indexes using an F-ratio test.
```{r}
#use same degrees of freedom calculated above.
#calculate F-ratio and critical value for s'max, again paying attention to whether males or females have the greater s'max value
###note: Jones index is in units of SD, so had to convert to variances for F ratio test
if((Jones_f^2) > (Jones_m^2)){ #if squared s'max is greater than squared male s'max
  
  crit <- qf(0.95, df1 = df_f, df2 = df_m) # use female df as df1 in calculation of critical value
  f <- (Jones_f^2)/(Jones_m^2) #use female estimate (squared) as numerator in f ratio
   pvalue <- pf(f, df1 = df_f, df2 = df_m, lower.tail = FALSE) #calculate p-value
  cat("Note: Female s'max (squared) value is numerator.\n\n") #cat description
  
} else { #otherwise
  
  crit <- qf(0.95, df1 = df_m, df2 = df_f) # use male df as df1 in calculation of critical value
  f <- (Jones_m^2)/(Jones_f^2) #use male estimate (squared) as numerator in f ratio
   pvalue <- pf(f, df1 = df_m, df2 = df_f, lower.tail = FALSE) #calculate p-value
  cat("Note: Male s'max (squared) value is numerator.\n\n") #cat description
  
}

#test whether male and female s'max are significantly different by comparing F ratio to critical value
if(f > crit){ #if the f ratio is larger than the critical value
  
  cat("Mean male and female s'max are different (F > critical).", "\n", "Male s'max: ", round(Jones_m, digits = 3), "\n", "Female s'max: ", round(Jones_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom: ", df_f, "\n", "P-value =", round(pvalue, digits = 3), "\n\n\n") #cat statistical metrics and parameters
  
} else { #otherwise
  
  cat("Mean male and female s'max are not different (F < critical).", "\n", "Male s'max: ", round(Jones_m, digits = 3), "\n", "Female s'max: ", round(Jones_f, digits = 3), "\n", "F-Ratio: ", round(f, digits = 3), "\n", "Critical Value: ", round(crit, digits = 3), "\n", "Male Degrees of Freedom: ", df_m, "\n", "Female Degrees of Freedom:", df_f,"\n", "P-value =", round(pvalue, digits = 3), "\n\n\n" ) #cat statistical metrics and parameters
  
} 
```
**There is a significant difference between the male and female Jones Index for copperheads.**